Stability of variational quantum dynamics with Gaussian approximants

### 2013-08-15 paper.tex
writing

@@

Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.  It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.  The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.  (Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.

@@

>>>>

%% Overview

There appear to be two sorts of instabilities.  The first, static, sort occurs when a quantum state is expanded over a finite set of coherent states, and those states have irregular amplitudes.  This is similar to the kind that numerical analysts call ill-posed problems, and the methods they use to analyse and resolve them can be applied here.  However, the cause of ill-posedness in our problem differs from those usually studied, and the usual methods of resolution don't work as well as they usually do.  There is one advantage of the expansion problem: because it is iterated at each timestep, the dynamics of the amplitudes can be exploited to so that they posed better expansion problems as the simulation progresses.

The second type of instability is a dynamical one.  This occurs whenever Schrödinger's equation is integrated with a quartic hamiltonian, over discrete timesteps.  In the second quantised picture, it occurs because the repulsion between particles causes components where many particles occupy the mode to oscillate very rapidly.  The simulation becomes stable when these components are removed, but that changes the dynamics.  Integrating Schrödinger's equation with repulsive particles either requires very short timesteps, or special integration formulae that exploit features of Hamiltonian dynamics.  Applying these formulae to the variational dynamics of the coherent states, which are not Hamiltonian, is somewhat subtle.

@@

Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
$$〈 Γ⁺|ψ〉(α)=π^{-½}e^{½|α|²}〈 α|ψ〉.$$
Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.  It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.  The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.  (Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.

@@

\beginsection{Convergence of discretely sampled expansions}

To get a sense of what residuals to expect from optimised expansions, the expansion of a state as an entire function, $f=〈Γ⁺|ψ〉$, can be sampled at discrete amplitudes, and the integral $|Γ〉f$ evaluated as a sum.  In the limit of many amplitudes, this should converge to $|Ψ〉$.  I chose the vacuum state as $|Ψ〉$.  I evaluated the integral in two ways, supposed to be extreme cases.

The first way used a grid, restricted to the disk $|α|<2.5$.  The pitch of the grid, $h$, was varied, and $f(α_r)=π^{-½}e^{½|α_r|²}〈α_r|ψ〉$ evaluated at each point.  The integral was discretised in the obvious way, to give approximations
$$|ψ_R〉={1\over πh²} ∑_{r=1}^R |α_r〉〈α_r|ψ〉.$$
A residual $‖|ψ〉-|ψ_R〉‖$ was found for each $R$; these are plotted as the solid line in the figure.  As the components increase, the residual asymptotes to the phase space truncation error $‖|Γ_χ〉〈Γ⁺|ψ〉-|ψ〉‖$.  (Calculate that precisely.)  The gaussian function $f$ is very smooth, so this converges rapidly as $R$ increases.

The second way was a type of stochastic integration.  The samples $α_r$ were drawn randomly, from the normal distribution $|〈α|ψ〉|$.  The integral was approximated by 
$$|ψ'_R〉=∑_{r=1}^R |α_r〉{〈α_r|ψ〉\over |〈α_r|ψ〉|},\qquad |ψ_R〉={|ψ'_R〉\over||ψ'_R〉|}.$$
The phase of $f$ is integrated by sampling, but the modulus is integrated stochastically.  The residuals are plotted as the broken line.  In principle, this will converge exactly as $R$ increases.  However, the convergence is $R^{-½}$, as seen on the graph, and expected from the laws of large numbers.

These two lines set bounds on how well we can reasonably expect to solve the expansion problem on an irregular grid.  A least squares method that does no better than sampling the argument of $f(α)$ at random $α$ is worthless: we need to do better than the dashed line.  On the other hand, it seems unreasonable to expect samples on an irregular grid to fit $|Ψ〉$ better than samples on a square grid.  If we try to match the solid line, we should expect the expansion coefficients to be very sensitive to the grid, and not to resemble $f(α_r)$.  If we try to do better than the truncation error, we will have to represent large number components, and the expansion coefficients will become very large.

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.
\endinsert
